---
title: \vspace{-1.5cm} MLM Assignment 2
author: 
- "Christine Hedde-von Westernhagen"
- "Emilia Löscher" 
- "Kyuri Park"
date: "*Utrecht University*, 09 March, 2022"
output: pdf_document
geometry: margin=0.7in
header-includes:
   - \usepackage{multirow,booktabs,setspace,caption,threeparttable}
   - \DeclareCaptionLabelSeparator*{spaced}{\\[2ex]}
   - \captionsetup[table]{textfont=it,format=plain,justification=raggedright, singlelinecheck=false,labelsep=spaced,skip=0pt}
   
---

\fontsize{10}{14}
\selectfont

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#(echo = TRUE, results = "hide")

library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)
library(pander)
library(purrr)
library(lattice)
library(gridExtra)

library(lme4)
library(lmerTest)
```

## Data Description

In this assignment, we analyze the `curran_wide.csv` data, which contains the information about the age, antisocial behavior, reading skills, emotional support, cognitive stimulation, and mother's age of 221 sampled children. Antisocial behavior and reading skills are measured over 4 occasions. In this analysis, we do not use the variables for child's gender and emotional support.

The (pre-processed) data specifics are as follows:

 * `id`: child id
 * `time`: measurement occasion ranging from 0 to 3 
 * `anti`: antisocial behavior (time-variant)
 * `read`: reading recognition skills (time-variant & grand-mean centered)
 * `momage`: mother's age measured at the first occasion (time-invariant & grand-mean centered)
 * `homecog`: cognitive stimulation measured at the first occasion (time-invariant & grand-mean centered)
 
 
## 1. Convert the wide data file into a long format. Check the data and recode if necessary.
As seen below, the original data (`curran_wide.csv`) is converted into a long format.  
`time` is recoded from 1 - 4 to 0 - 3 such that the first measurement is set to 0.  
`read`, `momage`, and `homecog` are centered at the grand mean such that the the intercept represents the expected value of antisocial behavior for an average child (with average reading skills + average mom's age + average cognitive stimulation) when these predictors are included in the model.  
***Maybe we should center each when adding them.. not sure about centering read if it makes any sense***

```{r, echo=FALSE, message=FALSE}
## load the data
curran_wide <- readr::read_csv("curran_wide.csv")
## check the wide data
head(curran_wide)
```

```{r, echo=FALSE}
## prepare the data
curran_long <- curran_wide %>% 
  # not using homeo and child's gender
  select(-c(homeemo,sex)) %>% 
  # convert it to a long format
  pivot_longer(!c(id, momage, homecog), 
               names_to = c(".value", "time"),
               names_pattern = "(anti|read)(.)",
               names_transform = list(time = as.integer))  %>% 
  # re-code time from 1:4 to 0:3, and grand-mean center 'read', 'momage' and 'homecog'
  mutate(time = time - 1,
         read = read - mean(read), 
         momage = momage - mean(momage), 
         homecog = homecog - mean(homecog)) %>% 
  # re-order the columns
  relocate(time, anti, read, .after=id)

## check the long formatted data
head(curran_long)

summed.dat <- psych::describe(curran_long)[,-c(1,6,7,10,12)] 
panderOptions('round',2)
pander(summed.dat, caption="Descriptive statistics")
```


```{r hist, echo=FALSE, fig.height =2, fig.width=5, fig.align='center', fig.cap="Right-skewed Antisocial behavior"}
# set the theme for the plots
My_Theme = theme(
  plot.title = element_text(size = 11, hjust = 0.5),
  axis.title.x = element_text(size = 10),
  axis.title.y = element_text(size = 10))

# plot the histogram of 'anti'
ggplot(curran_long, aes(x = anti)) +
  geom_histogram(binwidth=1, 
                 fill="#69b3a2", 
                 color="#e9ecef", 
                 alpha=0.9) +
  theme_minimal() +
  xlab("antisocial behavior") +
  ggtitle("Distribution of antisocial behavior") + 
  My_Theme
```

*Table 1* shows the descriptive statistics of each variable in the re-coded data.
There are total 884 obervations. The mean values of `read`, `momage`, and `homecog` are 0, as expected, and they seem to be not overly skewed given that they have small skewness (i.e., <|1|). The skewness value of `anti` is higher than 1, which indicates that the distribution is skewed. Hence, we checked on the distribution of `anti` and as shown in *Figure 1*, it is right-skewed where most values are clustered around the left tail. This is also aligned with the (top-left) boxplot shown in *Figure 2*.


### 1a. Check the linearity assumption, report and include plots.
Based on *Figure 3*, it is concluded that the linearity assumption is met in both level 1 and level 2, even though the relationships between the variables seem to be rather weak (i.e., slopes are quite flat), except for the one *(d) Cognitive stimulation - Average Antisocial behavior*.  

### 1b. Check for outliers.
*Figure 2* shows that *antisocial behavior* has several univariate outliers and *cognitive stimulation* also has one outlier in the left tail. In *Figure 3*, we can check the bivariate outliers at each level. At level 1, although there are few points spotted in the upper-side of plots *(a)* and *(b)*, they do not seem to be influential. Also, at level 2, no such influential outlier is observed. 


```{r boxplots, echo=FALSE, fig.height=3, fig.width=8, fig.align='center', fig.cap="Overall distribution of each variables"}

box_list <- list()

box_list <- map2(
  
  c("anti", "read", "momage", "homecog"),
  c("antisocial behavior", "reading recognition", 
       "mother's age", "cognitive stimulation"),
  
  function(var, lab) {
    ggplot(curran_long, aes(y = .data[[var]])) + 
      geom_boxplot() + 
      theme_minimal() + 
      labs(title = lab, y = "") + 
      coord_flip() +
      My_Theme
     }
  )

ggarrange(plotlist = box_list, nrow = 2, ncol = 2)
```

```{r scatterplots, echo=FALSE, fig.height=6, fig.width=8, message = FALSE, fig.cap="Scatterplots to insepct linearity and outliers"}
## check for linearity and outliers

## level1 
p1 <- ggplot(curran_long,
       aes(x = time, y = anti)) +
  geom_jitter(shape = 1, cex = 0.5) +
  geom_smooth(method = "lm",
              aes(color = "linear"),
              se = FALSE) +
  geom_smooth(method = "lm",
              formula = y ~ x + I(x^2),
              aes(color = "quadratic"),
              se = FALSE) +
  theme_minimal() +
  xlab("Time") +
  ylab("Antisocial behavior") +
  labs(color = "Fit method") +
  ggtitle("(a) Level 1 (occasion level)")+
  My_Theme

p2 <- ggplot(curran_long,
       aes(x = read, y = anti)) +
  geom_point(shape = 1, cex = 0.5) +
  geom_smooth(method = "lm",
              aes(color = "linear"),
              se = FALSE) +
  geom_smooth(method = "lm",
              formula = y ~ x + I(x^2),
              aes(color = "quadratic"),
              se = FALSE) +
  theme_minimal() +
  xlab("Reading recognition") +
  ylab("Antisocial behavior") +
  labs(color = "Fit method") +
  ggtitle("(b) Level 1 (occasion level)") + 
  My_Theme

## level2 
p3 <- curran_long %>% 
  group_by(id) %>% 
  # aggregate the antisocial behavior: aggregated_anti
  mutate(aggregated_anti = mean(anti)) %>% 
  ggplot(aes(x = momage, y = aggregated_anti)) +
    geom_point(shape = 1, cex = 0.5) +
  geom_smooth(method = "lm",
              aes(color = "linear"),
              se = FALSE) +
  geom_smooth(method = "lm",
              formula = y ~ x + I(x^2),
              aes(color = "quadratic"),
              se = FALSE) +
  theme_minimal() +
  xlab("Mother's age") +
  ylab("Avg. antisocial behavior over time") +
  labs(color = "Fit method") +
  ggtitle("(c) Level 2 (person level)") + 
  My_Theme

p4 <- curran_long %>% 
  group_by(id) %>% 
  # aggregate the antisocial behavior: aggregated_anti
  mutate(aggregated_anti = mean(anti)) %>% 
  ggplot(aes(x = homecog, y = aggregated_anti)) +
    geom_point(shape = 1, cex = 0.5) +
  geom_smooth(method = "lm",
              aes(color = "linear"),
              se = FALSE) +
  geom_smooth(method = "lm",
              formula = y ~ x + I(x^2),
              aes(color = "quadratic"),
              se = FALSE) +
  theme_minimal() +
  xlab("Cognitivie stimulation") +
  ylab("Avg. antisocial behavior over time") +
  labs(color = "Fit method") +
  ggtitle("(d) Level 2 (person level)") + 
  My_Theme

ggarrange(p1, p2, p3, p4, common.legend = TRUE, legend = "bottom")
```

## 2. Answer the question: should you perform a multilevel analysis?

### 2a. What is the mixed model equation?

 -- Mixed Model Equation
 
$$ y_{ti} = \beta_{00} + u_{0i} + e_{ti} $$

\begingroup
\fontsize{9}{12}\selectfont
\begin{itemize}
\setlength{\itemindent}{0.2cm}
\setlength{\itemsep}{0pt}
\setlength{\parskip}{0pt}
\setlength{\parsep}{0pt}
 \item[-] $y_{ti}$ refers to antisocial behavior of child $i$ at time $t$.
 \item[-] $\beta_{00}$ refers to the overall intercept, which is the average antisocial behavior over all children.
\item[-] $u_{0i}$ refers to the random residual error at the person level (level 2), which represents the deviation from the overall intercept ($\beta_{00}$) of child $i$. 
\item[-] $e_{ti}$ refers to the residual error at the occasion level (level 1).
\end{itemize}
\endgroup

### 2b. Provide and interpret the relevant results.
As shown in *Table 2*, the intercept term is identical as 1.82 in $M_{0}$ and $M_{1}$. This overall intercept represents the average antisocial behavior across all children. $M_{1}$ decomposes the variance term in a variance at level 1 ($Var_{occ}$) and level 2 ($Var_{sub}$).

The deviance for the random intercept model turns out to be significantly smaller than that of the single-level model, $\chi^2(1) = 231.97$, $p < .001$, indicating that $Var_{occ}$ is significantly greater than 0. AIC value is also lower with the random intercept model ($AIC_{M_{1}} = 3343.5$) compared to the single-level model ($AIC_{M_{0}} = 3573.5$), which is in accordance with the deviance difference test result. 

```{r, results='hide'}
# model0: single-level regression model for comparison
model0 <- lm(anti ~ 1, data = curran_long)
summary(model0)

# model1: random intercept model ((benchmark model to compute ICC))
model1 <- lmer(anti ~ 1 + (1|id), REML = FALSE, data = curran_long)
summary(model1)

# check the significance of random intercept
anova(model1, model0)
```


\begin{table}[!ht]
\centering
\begin{threeparttable}
\caption{Single-level model and Intercept-only model}
\begin{tabular}{lcc}
\toprule
\textbf{Model}                & $M_{0}$: single-level model & $M_{1}$: random intercept \\ \midrule
\textit{\textbf{Fixed part}}  & Coefficient(SE)             & Coefficient(SE)          \\
Intercept                     & 1.82(.06)                  & 1.82(.10)              \\
\textit{\textbf{Random part}} &                             &                         \\
$Var_{occ}$                      & 3.32                        & 1.74       \\
$Var_{sub}$                     &                             & 1.58                    \\
\textbf{Deviance}             & 3569.5                     & 3357.5                 \\
\textbf{AIC}                  & 3573.5                     & 3343.5                 \\
\textbf{Deviance difference$^{a}$}  &                             & 231.97$^{***}$                \\
\bottomrule
\multicolumn{3}{l}{Note}\\
$^{***} p<.001, ^{**} p<.01$ \\
$^a$ p-value for $\chi^2$ test is one-sided p-value. \\
\end{tabular}
\end{threeparttable}
\end{table}



### 2c. What is the intraclass correlation?

The intraclass correlation ($\rho$) is calculated as follows: 

$$ \rho = \frac{\sigma^2_{u0}}{\sigma^2_{u0} + \sigma^2_e} $$ 

```{r}
ICC <- 1.579/(1.579+1.741)
```

The ICC in this model is equal to $\rho = `r round(ICC, digits=2)`$.

### 2d. What is your conclusion regarding the overall question regarding the necessity of performing a multilevel analysis?  

Yes, we should perform the multilevel analysis in this case, because not only the data structure is nested (i.e., multiple measurements within each individual), but also the difference between individuals accounts for about *48%* of the total variance. In other words, the intraclass correlation -- ICC: the proportion of the total variance explained by the between-individual differences -- is *0.476*, which is high enough that the multilevel analysis is warranted.

In addition, *Figure 4* shows the relationships between *antisocial behavior - time* and *antisocial behavior - reading skills* in each children (first 20). Here we see that the slopes and intercepts vary across the children, which again suggests the necessity of multilevel analysis that can properly address this between-person variabilities.

```{r schoolscatter, echo=FALSE, fig.align='center', fig.cap="Scatterplots of children to inspect the variabilities in intercept/slopes", fig.height=10}
uniqueID <- unique(curran_long$id)[1:20]
p5 <- xyplot(anti ~ time | id, groups= id, data=curran_long[curran_long$id %in% uniqueID,],
       type=c('p','r'), auto.key=F, cex=0.3, main= "(a) Antisocial behavior - Time")

p6 <- xyplot(anti ~ read | id, groups= id, data=curran_long[curran_long$id %in% uniqueID,],
       type=c('p','r'), auto.key=F, cex=0.3, main= "(b) Antisocial behavior - Reading skills")

grid.arrange(p5, p6)
```

## 3. Add the time-varying predictor(s).
### Provide and interpret the relevant results and provide your overall conclusion.

***- mention centering technique <-  I forgot what this comment was about exactly ....***

As shown in *Table 3*, the intercept term is now 1.55 in $M_2$ with `time` predictor, which refers to the average antisocial behavior at the first occasion (time = 0). `time` turns out to be a statistically significant predictor of `anti`, $b_{time}=0.18, t(663) = 4.513, p <.001$, meaning that antisocial behavior is expected to increase by 0.18 on each succeeding occasion. By adding `time`, the occasion-level residual error variance ($Var_{occ}$) goes down to 1.69, but the subject-level variance ($Var_{sub}$) goes up to 1.60. This is because in the longitudinal data with fixed occasion measurements, there is no variation in time-points between subjects. Subsequently, $Var_{occ}$ is overestimated and $Var_{sub}$ is underestimated in the intercept-only model, $M_1$. In order to correct this, measurement occasion variable needs to be included in the model. Hence, $M_2$ with the `time` predictor correctly decomposes the variance terms and correspondingly can be used as a benchmark model for computing the $R^2$.  

The significance of adding `time` to the random intercept model($M_{1}$):  
The deviance of $M_{2}$ is significantly smaller than the deviance of $M_{1}$, $\chi^2(1) = 20.06$, $p < .001$ (see *Table 2* and *Table 3*). It indicates that $M_{2}$ fits significantly better than the intercept-only model, $M_{1}$. AIC also corresponds to this, as $AIC_{M_{2}} = 3325.5$ is lower than $AIC_{M_{1}} = 3343.5$.

In $M_{3}$ with `time` and `read` together, the `time` is still a significant predictor $b_{time}=0.21, t(882) = 2.73, p <.01$, but `read` is not a significant predictor of `anti`, $b_{read}=-0.03, t(831) = -0.54, p = .588$. The deviance of $M_3$ is 3317.2, which is almost the same as $M_2$, and the difference in the deviances is not significant as expected, $\chi^2(1) = 0.29$, $p = .297$. Also, AIC of $M_3$ is slightly higher than AIC of $M_2$ (see *Table 3*), which again suggests that adding `read` does not provide a better fit. Therefore, we drop `read` and proceed with $M_2$ including only the `time` predictor.

```{r results='hide'}
# model2: add a time-varying predictor, time ((benchmark model for computing R2))
model2 <- lmer(anti ~ 1 + time + (1|id), REML = FALSE, data = curran_long)
summary(model2)
anova(model2, model1)

# model3: add a time-varying predictor, read
model3 <- lmer(anti ~ 1 + time + read + (1|id), REML = FALSE, data = curran_long)
summary(model3)
anova(model3, model2)
```


\begin{table}[!ht]
\centering
\begin{threeparttable}
\caption{Adding time-varying predictors}
\begin{tabular}{lcc}
\toprule
\textbf{Model}                & $M_{2}$: add time & $M_{3}$: add read \\ \midrule
\textit{\textbf{Fixed part}}  & Coefficient(SE)             & Coefficient(SE)          \\
Intercept                     & 1.55(.11)                   & 1.50(.15)              \\
time                          & 0.18$(.04)^{***}$           & 0.21$(.07)^{**}$     \\
read                          &                             & -0.03$(.06)$           \\
\textit{\textbf{Random part}} &                             &                         \\
$Var_{occ}$                   & 1.69                        & 1.69       \\
$Var_{sub}$                   & 1.60                        & 1.58                    \\
\textbf{Deviance}             & 3317.5                      & 3317.2                 \\
\textbf{AIC}                  & 3325.5                      & 3327.2                 \\
\textbf{Deviance difference$^{a}$}  & 20.06$^{***}$         & 0.29               \\
\bottomrule
\multicolumn{3}{l}{Note}\\
$^{***} p<.001, ^{**} p<.01$ \\
$^a$ p-value for $\chi^2$ test is one-sided p-value. \\
\end{tabular}
\end{threeparttable}
\end{table}


## 4. On which level or levels can you expect explained variance?
### Calculate and interpret the explained variances.
In theory, we can expect the explained variances ($R^2$) at both occasion (level 1) and subject level (level 2), as the level 1 predictor in general can explain the variances in both levels. However, in this case where the additional time-varying predictor `read` is not significant and correspondingly does not reduce  the variance at the occasion level (but rather increase it slightly), $R^2_{occasion}$ becomes negative and therefore cannot be defined.  

```{r, results='hide'}
m2var.lv1 <- 1.689 # level 1 variance in model 2 (benchmark model)
m2var.lv2 <- 1.592 # level 2 variance in model 2 (benchmark model)
m3var.lv1 <- 1.693 # level 1 variance in model 3 
m3var.lv2 <- 1.576 # level 2 variance in model 3

# explained variance at level 1 (occasion level)
R2.lv1 <- (m2var.lv1 - m3var.lv1) / m2var.lv1
R2.lv1

# explained variance at level 2 (subject level)
R2.lv2 <- (m2var.lv2 - m3var.lv2) / m2var.lv2
R2.lv2
```

The computed $R^2$ values for each level using $M_2$ as our benchmark:  

 - $R^2_{occasion}$ = not defined
 - $R^2_{subject}$ = $`r round(R2.lv2, 4)`$

It means that reading recognition skill (`read`) further explains about 1% of the variance between children.


## 5. Add the time invariant predictor(s) to the model.
### Provide and interpret the relevant results and provide your overall conclusion.
In $M_{4a}$ we add both time invariant predictors, mother's age (`momage`) and cognitive stimulation (`homecog`). As shown in *Table 4*, `momage` is turned out to be not significant, $b_{momage}=-0.001, t(221) = -0.02, p = .985$, and only `homecog` is significant, $b_{homecog}=-0.13, t(221) = -3.35, p < .001$. This means that for each one point increase in `homecog` (the more cognitive stimulation a child receives), the average `anti` is expected to decrease by 0.13 (the average antisocial behavior of a child goes down). The intercept term value remains almost the same as the previous model (1.55), but the interpretation now with the added predictors is the average antisocial behavior for a child with the average cognitive stimulation and average mom's age at the first occasion (as `homecog` and `momage` are centered at the grand mean).  
By adding these level 2 predictors, we see that the variance at level 2 (subject level) is decreased and this correspondingly produces $R^2_{subject}=.065$.  
The significance of adding `momage` and `homecog` to $M_{2}$:  
The deviance of $M_{4a}$ is significantly smaller than the deviance of $M_{2}$, $\chi^2(2) = 11.642$, $p < .01$. It indicates that $M_{4a}$ fits significantly better than $M_{2}$ with only `time`. 
The lower $AIC_{M_{4a}} = 3317.8$ also suggests that $M_{4a}$ is preferable.  

However, given that `momage` is not a significant predictor, we drop `momage` and only include `homecog` in the model, $M_{4b}$. The fixed part of model does not change much, both `time`, $b_{time}=0.18, t(663) = 4.513, p < .001$, and `homecog` are significant,$b_{homecog}=-0.13, t(221) = -3.35, p < .001$. The interpretation of `homecog` is the same as described above, but the intercept term would now represent the average antisocial behavior for a child with the average cognitive stimulation at the first occasion.  
By keeping only `homecog` in the model, we see that the variance at level 2 (subject level) is decreased by the same amount as in $M_{4a}$ and accordingly results in the same $R^2_{subject}=.065$.  
The significance of adding only `homecog` to $M_{2}$:  
The deviance of $M_{4b}$ is significantly smaller than the deviance of $M_{2}$, $\chi^2(1) = 11.641$, $p < .001$. It indicates that $M_{4b}$ fits significantly better than $M_{2}$ with only `time`. 
In addition, $AIC_{M_{4b}} = 3315.8$, which is not only lower than $M_{2}$ but also slightly lower than $M_{4a}$, indicating that M_{4b} is a better fit than M_{4a}. Given these overall results, we decided to proceed with M_{4b} with `time` and `homecog` predictors.

```{r, results='hide'}
## proceed with a model without the non-significant predictor, 'read'
# model4a: add time-invariant predictors, momage & homecog
model4a <- lmer(anti ~ 1 + time + momage + homecog + (1|id), REML = FALSE, data= curran_long)
summary(model4a)
anova(model4a, model2)

# model4b: remove the non-significant predictor, 'momage'
model4b <- lmer(anti ~ 1 + time + homecog + (1|id), REML = FALSE, data= curran_long)
summary(model4b)
anova(model4b, model2)
```

\begin{table}[!ht]
\centering
\begin{threeparttable}
\caption{Adding time-invariant predictors}
\begin{tabular}{lcc}
\toprule
\textbf{Model}                  & $M_{4a}$: add momage + homecog & $M_{4b}$: add only homecog \\ \midrule
\textit{\textbf{Fixed part}}      & Coefficient(SE)                & Coefficient(SE)          \\
Intercept                         & 1.55(.11)                      & 1.55(.11)              \\
time                              & 0.18$(.04)^{***}$               & 0.18$(.04)^{***}$     \\
momage                            & 0.00(.05)                       &                       \\
homecog                           & -0.13$(.04)^{***}$               & -0.13$(.01)^{***}$     \\
\textit{\textbf{Random part}}     &                                 &                         \\
$Var_{occ}$                       & 1.70                            & 1.69       \\
$Var_{sub}$                       & 1.50                            & 1.49                    \\
\textbf{Deviance}                 & 3305.8                         & 3305.8                 \\
\textbf{AIC}                      & 3317.8                         & 3315.8                 \\
\textbf{Deviance difference$^{ab}$}  &11.64$^{**}$            & 11.64$^{***}$               \\
\bottomrule
\multicolumn{3}{l}{Note}\\
$^{***} p<.001, ^{**} p<.01$ \\
$^a$ p-value for $\chi^2$ test is one-sided p-value. \\
$^b$ Here the deviance is compared with $M_{2}$.
\end{tabular}
\end{threeparttable}
\end{table}

## 6. On which level or levels can you expect explained variance?
### Calculate and interpret the explained variances.
We can expect the explained variances ($R^2$) at the subject level (level 2), as the level 2 predictor can only explain the variance in level 2. 

```{r, results='hide'}
m2var.lv1 <- 1.689 # level 1 variance in model 2 (benchmark model)
m2var.lv2 <- 1.592 # level 2 variance in model 2 (benchmark model)
m4var.lv1 <- 1.689 # level 1 variance in model 4b    
m4var.lv2 <- 1.488 # level 2 variance in model 4b

# explained variance at level 2 (subject level)
R2.lv2 <- (m2var.lv2 - m4var.lv2) / m2var.lv2
R2.lv2
```

The computed $R^2$ value for the subject level using $M_2$ as our benchmark:   

 - $R^2_{subject}$ = $`r round(R2.lv2, 4)`$  
 
It means that cognitive stimulation provided at home (`homecog`) explains about 6% of the variance between children.

## 7. For the time-varying predictor(s), check if the slope is fixed or random.

### 7a. What are the null- and alternative hypotheses?

 - *$H_{0}$*: $\sigma_{u1}^2 = 0$; The slope for the time variable is equal across the children. 
 - *$H_{1}$*:  $\sigma_{u1}^2 > 0$; The slope for the time variable varies across the children. 

### 7b. Provide and interpret the relevant results.
- interpret result chi2 test 

```{r, results='hide'}
# model5: let 'time' have random slopes
model5 <- lmer(anti ~ 1 + time + homecog + (1 + time|id), REML = FALSE, data = curran_long)
summary(model5)
anova(model5, model4b)

# # model5b: let 'read' have random slopes
# model5b <- lmer(anti ~ 1 + time + homecog + (1 + read|id),
#                 REML = FALSE, data= curran_long)
# summary(model5b)
# 
# model5b2 <- lmer(anti ~ 1 + time + homecog + read + (1 + read|id),
#                 REML = FALSE, data= curran_long)
# summary(model5b2)
# 
# anova(model5b, model4)
# 
# # model5c: let both have random slopes --> model fails to converge
# model5c <-  lmer(anti ~ 1 + time + homecog + (1 + time + read|id), 
#                  REML = FALSE, data = curran_long)
```
\begin{table}[!ht]
\centering
\begin{threeparttable}
\caption{Model with random slope and covariance of intercept and slope}
\begin{tabular}{lc}
\toprule
\textbf{Model}                & $M_{5}$: random slope \\ 
\midrule
\textit{\textbf{Fixed part}}  & Coefficient(SE)       \\
Intercept                     & 1.55 (.10)            \\
time                          & 0.18$(.04)^{***}$             \\
homecog                       &-0.10$(.04)^{**}$            \\
\textit{\textbf{Random part}} &                       \\
$Var_{occ}$                & 0.95                  \\
$Var_{sub}$               & 1.53                  \\
$Var_{time}$               & 0.10                  \\
$Cor_{sub*time}$                & 0.41                  \\
\textbf{Deviance}             & 3279.3                \\
\textbf{AIC}                  & 3293.3                \\
\textbf{Deviance difference$^{ab}$}  &26.56$^{***}$  \\
\bottomrule
\multicolumn{2}{l}{Note}\\
$^{***} p<.001, ^{**} p<.01$ \\
$^a$ p-value for $\chi^2$ test is one-sided p-value.\\
$^b$ Here the deviance is compared with $M_{4b}$.
\end{tabular}
\end{threeparttable}
\end{table}


### 7c. Provide an overall conclusion.

The variance of time is significant when added to the model therefore we conclude that the effect of...


## 8. If there is a random slope, set up a model that predicts the slope variation.
We have found that there is a random slope for the variable `time`.
In the following, we are fitting two models to check whether the variable `momage` or the 
variable `homecog` can predict the slope variation.

### Provide and interpret the relevant results and provide your overall conclusion.
From model 6a, it can be seen that `homecog` can predict the slope variation of `time`
as the interaction of `homecog` and `time` is significant ($p$ = 0.008).
Note, however, that the fixed effect for `homecog` is not significant anymore after
including the interaction term ($p$ = 0.1056). Nevertheless, the fixed effect of
`homecog` is not removed from the models as fixed effect that are included part of an
interaction should always remain in the model.
The estimate for the interaction term is -0.04532. This means that for children 
who have more cognitive stimulation (higher values for `homecog`),
their antisocial behavior increases less strongly over time. This is visualized 
in the figure below. 

In model 6b the interaction of `time` and `momage` was added to the model. 
From the summary of model 6b, it can be seen that `momage` cannot be used to
predict the slope variation of `time` as the interaction of `momage` and `time` 
is not significant ($p$ = 0.88749).
Furthermore, the fixed effect of `momage` is also not significant ($p$ = 0.90528).
This had already been the case in model 4a and, therefore, `momage` had been removed 
from the model. It was added here once more because it is part of the interaction.

```{r, results='hide'}
## model6: add a cross-level interaction
# check if homecog can explain the time variance
model6a <- lmer(anti ~ 1 + time + homecog + homecog*time + (1+time|id), 
                REML = FALSE, data = curran_long)
summary(model6a)
anova(model6a, model5)

# check if momage can explain the time variance
model6b <- lmer(anti ~ 1 + time + momage + homecog + momage*time + (1+time|id), 
                REML = FALSE, data = curran_long)
summary(model6b)
anova(model6b, model5)
```

\begin{table}[!ht]
\centering
\begin{threeparttable}
\caption{Adding a cross-level interaction}
\begin{tabular}{lcc}
\toprule
\textbf{Model}                  & $M_{6a}$: `homecog*time` & $M_{6b}$: `momage*time` \\ \midrule
\textit{\textbf{Fixed part}}      & Coefficient(SE)                & Coefficient(SE)          \\
Intercept                         & 1.55(.10)                      & 1.55(.10)              \\
time                              & 0.18$(.04)^{***}$               & 0.18$(.04)^{***}$     \\
momage                            &                                  &-0.01(.05)                       \\
homecog                           & -0.06(.04)                       & -0.10$(.04)^{**}$     \\
time:homecog                      & -0.05$(.02)^{**}$               &           \\
time:momage                      &                                  &0.00(.02)       \\
\textit{\textbf{Random part}}     &                                 &                         \\
$Var_{occ}$                       & 1.53                            & 1.53       \\
$Var_{sub}$                       & 0.94                            & 0.95                    \\
$Var_{time}$                     & 0.08                             &    \\
$Cor_{sub*time}$                & 0.47                              & 0.41   \\
\textbf{Deviance}                 & 3272.4                         & 3279.2                 \\
\textbf{AIC}                      & 3288.4                         & 3297.2                 \\
\textbf{Deviance difference$^{ab}$}  &6.88$^{**}$            & 0.05               \\
\bottomrule
\multicolumn{3}{l}{Note}\\
$^{***} p<.001, ^{**} p<.01$ \\
$^a$ p-value for $\chi^2$ test is one-sided p-value. \\
$^b$ Here the deviance is compared with $M_{5}$.
\end{tabular}
\end{threeparttable}
\end{table}

```{r, echo=FALSE, message=FALSE}
curran_long$predict <- predict(model6b)

homecog_labs <- as.character(sort(unique(round(curran_long$homecog, 2))))
names(homecog_labs) <- as.character(sort(unique(curran_long$homecog)))

ggplot(aes(time, predict), data = curran_long) +
  geom_jitter(size = .7, alpha = .7, width = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ homecog, 
             labeller = labeller(homecog = homecog_labs)) +
  ggtitle("Predicted values of antisocial behavior over time by levels of 'homecog'") +
  theme_minimal() +
  My_Theme
```

-> for children with more homecog, antisocial behavior increases less strongly/decreases over time

So `model6a` is our final model.

## 9. Decide on a final model.
We choose model 6a to be our final model as it has the best model fit among all 
the tested models. The level 1 and 2 model equations, as well as the mixed model 
equation are given below. 

### 9a. Provide the separate level 1 and 2 model equations, as well as the mixed model equation.


 -- Level 1 Model Equation
 
 $$anti_{ti} = \pi_{0i} + \pi_{1i}time_{ti} + e_{ti}$$
 
 -- Level 2 Model Equations
 
$$\pi_{0i} = \beta_{00} + \beta_{01}homecog_{i} + u_{0i} $$

$$\pi_{1i} = \beta_{10} + \beta_{11}homecog_{i} + u_{1i} $$
  
 -- Mixed Model Equation
 
$$ anti_{ti} = \beta_{00} + \beta_{10} time_{ti} + \beta_{01}homecog_{i} + \beta_{11}homecog_{i} \times time_{ti} + u_{0i} + u_{1i}time_{ti} + e_{ti} $$

### 9b. Check the normality assumption for both the level-1 and level-2 errors, report.

```{r}
# level 1 residuals
resid_lvl1 <- residuals(model6a)

# level 2 residuals
resid_lvl2 <- ranef(model6a)$id
```

```{r qqplots, echo=FALSE, fig.height=9, fig.width=9, fig.align='center', fig.cap="Q-Q plots for level 1 and level 2 residuals"}
# Set plot layout
layout(mat = matrix(c(1, 3, 2, 4), nrow = 2, ncol = 2),
       heights = c(2, 2), widths = c(2, 2))

# residual variance: e_ij
qqnorm(resid_lvl1, main = "(a) Normal QQ plot for Level 1 Residual Variance")
qqline(resid_lvl1, col = "salmon")

# fitted vs resid(e_ij) plot: check heteroscedasticity
plot(fitted(model6a), residuals(model6a), main = "(b) Residuals vs Fitted", 
     xlab="Fitted vales", ylab="Residuals")
abline(h=0, col="salmon")

# intercept variance: u_0j
qqnorm(resid_lvl2[,1], main = "(c) Normal QQ plot for Level 2 Intercept Variance")
qqline(resid_lvl2[,1], col = "salmon")

# slope variance: u_1j
qqnorm(resid_lvl2[,2], main = "(d) Normal QQ plot for Level 2 Slope Variance")
qqline(resid_lvl2[,2], col = "salmon")
```

- normality violated, anti very skewed see figure 1, consider log transformation (mention that we checked it)

We checked the normality assumption on level 1 and on level 2.
In all three cases, the normality assumption is (somewhat) violated. 
This can be seen from the respecitve Q-Q plots as the points were not on a straight diagonal line. 
They were either s-shaped for the level 1 errors or u-shaped for both of the level 
2 errors. 

The histogram of the outcome variable antisocial behavior (anti) showed that
the distribution of the variable is very (right) skewed (see Figure 1).
A log-transformation of the variable anti should be considered because this might
make the normality assumption more justifiable. When looking at the Q-Q plots
for model 6a using the log-transformed outcome variable anti,
the points in all three of the Q-Q plots were closer to being on a straight 
diagonal line than for the non-transformed variable. 
Hence, it might be a good idea to analyze the data with a log-transformed version
of the outcome variable. 

## Contribution

- Christine: Coding, interpretation, writing
- Emilia: Coding, interpretation, writing, Excel
- Kyuri: Coding, interpretation, writing

